Drug sensitivity prediction with the normal inverse Gaussian prior informed by external data

### 2019-11-27 docs/biometrical_main.Rnw
added EMSE split for large and small betas to simulations_gdsc.R

\documentclass[bimj,fleqn]{w-art}
\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,epsfig,
rotating,appendix,titlesec,multicol,hyperref,verbatim,
bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
dsfont,parskip,xr,tikz,times,w-thm}
\usepackage[]{graphicx}
\usepackage[authoryear]{natbib}
\externaldocument[sm-]{supplement1}
\usetikzlibrary{bayesnet}

@@

  \title[Drug sensitivity prediction with NIG prior]
  {Drug sensitivity prediction with the normal inverse Gaussian prior 
  informed by external data}

@@

	\subsection{Bayesian prior model}
	We carry out inference by endowing the parameters with the following priors:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), 
		  \label{eq:betaprior}\\
		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), \\
		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
		  \label{eq:tauprior}\\
		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
	with mean $\phi$ and shape $\lambda$. Prior distributions of the form
	(\ref{eq:prior}) are often referred to as global-local shrinkage rules 
	\cite[]{bernardo_shrink_2011}, due to the multiplicative separation of the 
	prior 
	variance into a local components $\gamma_{jd}^2$ and a global component
	$\tau_d^2$. The local components are supposed to capture local variation in
	the model parameters $\beta_{jd}$, while the global component is supposed to
	capture the general trend in $\beta_{jd}$. Addionally, for appropriate local
	shrinkage in global-local 
	shrinkage models it is important  to account for 
  different noise levels $\sigma_d^2$ by scaling the $\beta_{jd}$ variances 
  accordingly.

	The normal inverse Gaussian (NIG) prior model was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and since 
  \cite{barndorff-nielsen_normal_1997} it is routinely applied in mathematical 
  finance (see, e.g., \cite{kalemanova_normal_2007}). Here we extend it with an
  additional global variance component $\tau_d^2$. Supplementary Material
  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
  To illustrate the effect of the NIG prior on the posterior mean, we consider 
  the prior reparametrised as in 
  \cite{carvalho_handling_2009}, i.e., in terms of shrinkage weights 
  $\kappa_{jd}=1/(1 + \gamma_{jd}^2) \in (0, 1)$.
  Under the (simplified) normal means model, i.e., 
  $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
  $\tau_d^2=\sigma_d^2=1$, the resulting
  posterior mean for the $\beta_{jd}$ is
  $\E(\beta_{jd} | y_{jd}, \kappa_{jd}) = (1 - \kappa_{jd}) y_{jd}$. 
  Thus, $\kappa_{jd}=0$ implies no shrinkage of $\beta_{jd}$ and $\kappa_{jd}=1$
  implies full shrinkage towards zero. Figure 
  \ref{fig:dens_kappa} depicts the prior on $\kappa_{jd}$ implied by 
  several choices of $\beta_{jd}$ prior.
<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors. Different line types correspond to different hyperparameter settings. The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="90%", fig.asp=1/3>>=
@

@@

  \subsection{Empirical Bayes}\label{sec:empiricalbayes}
  We parametrised the prior 
  mean of the $\gamma_{jd}^2$ as $\phi_{jd}=(\mathbf{c}_{jd} \tr 
  \bm{\alpha}_{\text{feat}})^{-1}$ and the prior mean of $\tau_d^2$ as
  $\chi_{d}=(\mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}})^{-1}$.
  This parametrisation allows us to include feature and drug covariates,
  both continuous and discrete, into the 
  model. Additionally, it reduces the number of hyperparameters from 
  $pD$ to $|\bm{\alpha}_{\text{feat}}| + |\bm{\alpha}_{\text{drug}}| + 2$. The
  Bayesian model
  then requires the specification of the hyperparameters 
  $\bm{\alpha}=\begin{bmatrix} \bm{\alpha}_{\text{feat}} \tr & 
  \bm{\alpha}_{\text{drug}} \tr 
  \end{bmatrix} \tr$ and $\bm{\lambda} = \begin{bmatrix} \lambda_{\text{feat}} &
  \lambda_{\text{drug}}
  \end{bmatrix} \tr$. These are
  abstract and hard to interpret parameters 
  for which we generally lack expert knowledge. They do, however, have a 
  significant influence on the shape of the posterior distribution. We 
  therefore propose to estimate these hyperparameters by empirical Bayes.
  In our case, this results in an   
  objective and data-driven inclusion of the feature and drug covariates.

	The canonical method for empirical Bayes is to maximise the marginal 
	likelihood with respect to the hyperparameters. In 
	\cite{casella_empirical_2001} the marginal likelihood is maximised by an EM 
	algorithm:
	\begin{align*}
	  \bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} & = \underset{\bm{\alpha},
	  \bm{\lambda}}{\argmax}\E_{\cdot | \Y} 
	  [\log p(\Y, \B, \bm{\Gamma}^2, \bm{\tau}^2, \bsigma^2) | \bm{\alpha}^{(l)},
	  \bm{\lambda}^{(l)}] \\
	  & = \underset{\bm{\alpha},\bm{\lambda}}{\argmax} 
	  \E_{\cdot | \Y} [\log \pi (\bm{\Gamma}^2) | \bm{\alpha}_{\text{feat}}^{(l)},
	  \bm{\lambda}_{\text{feat}}^{(l)}] + 
	  \E_{\cdot | \Y} [\log \pi (\bm{\tau}^2) | 
	  \bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
	\end{align*}
	where $\Y = \begin{bmatrix} \y_1 & \cdots & \y_D \end{bmatrix}$,
	$\B = \begin{bmatrix} \bbeta_1, \dots, \bbeta_D \end{bmatrix}$, 
	$\bm{\tau}^2 = \begin{bmatrix} \tau^2_1 & \cdots & 
	\tau^2_D \end{bmatrix} \tr$,
	$\bsigma^2 = \begin{bmatrix} \sigma^2_1 & \cdots & 
	\sigma^2_D \end{bmatrix} \tr$,
	and $\bm{\Gamma}^2 = \begin{bmatrix} \bgamma^2_1 & \cdots & \bgamma^2_D 
	\end{bmatrix}$,
	and the expectation is with respect to the joint posterior. In our case, this 
	posterior is not available in closed form, which renders the expectation 
	difficult. While \cite{casella_empirical_2001} suggests to approximate the 
	expectation by a Monte Carlo sample, we propose to use the variational Bayes 
	approximation developed in Section \ref{sec:variationalbayes}:
	$$
	\bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} = 
	\underset{\bm{\alpha},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
	[\log \pi(\bm{\Gamma}^2)| \bm{\alpha}_{\text{feat}}^{(l)}, 
	\bm{\lambda}_{\text{feat}}^{(l)}] + 
	\E_{Q^{(l)}} [\log \pi(\bm{\tau}^2)| 
	\bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
	$$
	where now the expectation is with respect to the converged variational 
	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$.

@@

	\section{Simulations}
	\subsection{Fixed drug component}
	This section investigates the empirical 
	Bayes estimation properties of the model in a simulated setting; its main aim 
	is to assess hyperparameter recovery. It is a data based simulation,
	wherein the responses are simulated from a synthetic model, but the
	features are taken from the real GDSC data introducted in Section 
	\ref{sec:gdsc}. 
	The real GDSC features contain strong collinearities. Such strong
	collinearities in the design matrix impede correct parameter estimation. 
	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
	with the more modest aim of approximately correct estimation of the 
	hyperparameters.

	A pre-processing step, as explained in Section \ref{sec:gdsc}, selects around 
	300 features per drug, while the 168 drug sensitivities for 482 cell lines are 
	simulated from model (\ref{eq:likelihood}) and 
	(\ref{eq:prior}), with two adjustments: we fix $\forall d: \tau_d^2=1$ 
	and draw the error variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
	such that the prior $\sigma_d^2$ mean and variance are both one.
	We create four external dummy feature covariates that code for four 
	randomly assigned groups of features. We set $\bm{\alpha}_{\text{feat}}$ such
	that the $\gamma_{jd}^2$ of the four groups of features have prior means
	$\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$ 
	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. We estimate 
	four models: (i) the NIG model as in Section \ref{sec:estimation}, (ii) an 
	NIG model where we fix $\lambda_{\text{feat}}$ to its true value, and two NIG
	models without the inclusion of the external covariates, again with 
	$\lambda_{\text{feat}}$ either (iii) estimated or (iv) fixed to its true 
	value. Exclusion of the external covariates as in models (iii) and (iv) 
	amounts to direct estimation of one common expected 
	prior mean $\phi$, instead of regression estimates for the $\phi_{jd}$ as
	in models (i) and (ii). All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines.

<<include=FALSE>>=
load("results/simulations_gdsc_set1.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
melbo <- apply(elbo, 2, median)
phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
mphi <- apply(1/est[, c(11, 13)], 2, median)
mlambdaf <- apply(est[, c(5, 12)], 2, median)
@	
	Figure \ref{fig:simulations_gdsc_est1} shows the estimated
	$\bm{\alpha}_{\text{feat}}$ together with its true value for models (i) and 
	(ii). The figure shows that $\bm{\alpha}_{\text{feat}}$ is consistently
	underestimated 
	in the full NIG model (both with estimated and fixed $\lambda_{\text{feat}}$),
	while $\lambda_{\text{feat}}$ is overestimated as 
	\Sexpr{round(mlambdaf[1], 2)} in model (i) (the true value is 1). 
	This translates into
	underestimated prior means $\phi_{jd}$ as presented in Figure 
	\ref{fig:simulations_gdsc_phi1}, which in turn implies underestimated 
	$\beta_{jd}$ variances. The overestimated $\lambda_{\text{feat}}$ translates 
	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
	a common phenomenon in variational Bayes approximations. In addition, 
	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are higher level 
	parameters, which are, in general, hard to estimate. We note 
	that even though $\bm{\alpha}_{\text{feat}}$ and thus $\phi_{jd}$ are 
	estimated with bias, the orders of their elements are correct.
<<simulations_gdsc_est1, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>= 
@

<<simulations_gdsc_phi1, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ for the four groups of features from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>= 
@

	The models (iii) and (iv) give median $\phi$ estimates of
	\Sexpr{round(mphi[1], 2)} and \Sexpr{round(mphi[2], 2)}, 
	respectively. The $\phi$ estimates are
	underestimated, compared to true $\phi$ mean \Sexpr{round(mean(phi), 2)}. 
	Model (iii) slightly underestimates the median
	$\lambda_{\text{feat}}$ at \Sexpr{round(mlambdaf, 2)[2]} 
	(the true value is one) and thus overestimates the 
	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
	prior mean 
	$\phi$, that is not able to capture part of the variation in the true feature
	specific $\phi_{jd}$. The median evidence 
	lower bounds (ELBO), which approximate the marginal likelihoods of models
	(i)-(iv), are estimated as \Sexpr{round(melbo, 2)[1:3]}, and 
	\Sexpr{round(melbo, 2)[4]}, respectively. This implies that model (i),
	with estimated $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ fits
	the data best (the log Bayes factors comparing models (ii)-(iv) to model (i) 
	are \Sexpr{round(melbo[-c(1, 4)] - melbo[1], 2)}, and 
	\Sexpr{round(melbo[4] - melbo[1], 2)}, respectively, and thus highly favouring
	model (i)). The prediction MSEs for the four models on the test 
	data and gaps between training and test prediction MSE are displayed in Table
	\ref{tab:pmse1}. Table \ref{tab:pmse1} confirms that the models that
	include the external data (models (i) and (ii)) learn the underlying structure 
	in the data better than models (iii) and (iv) that do not include the external 
	data.
<<pmse1>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
load("results/simulations_gdsc_set1.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
options(knitr.kable.NA="", digits=3)
kable(tab,
      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
      caption="Median prediction MSEs for the models (i)-(iv) in the fixed 
      $\\tau_d^2$ simulation setting.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

  \subsection{Random drug component}
	To investigate the interplay of the feature and drug variance components, we 
	repeat the whole simulation study with random $\tau_d^2$. Similarly as with 
	the $\gamma_{jd}^2$, we randomly create four groups of drugs and corresponding
	external drug dummy covariates. We simulate the $\tau_d^2$ as in
	(\ref{eq:tauprior}), where we set $\bm{\alpha}_{\text{drug}} = 
	\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$, such
	that the $\tau_{d}^2$ of the four groups of features have prior means
	$\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The scale $\lambda_{\text{drug}}$ is 
	set to one. We estimate the same four models as in the fixed $\tau_d^2$ 
	setting, but in addition estimate the $\tau_d^2$ priors as well. Again, we
	use two different versions: one where we estimate 
	$\bm{\lambda}$ and one where we fix it to its true value.

>>>>

\documentclass[bimj,fleqn]{w-art}
\usepackage{bm,amsmath,amssymb,amsthm,amsfonts,graphics,graphicx,epsfig,
rotating,natbib,titlesec,multicol,hyperref,verbatim,
bbm,algorithm,algpseudocode,pgfplotstable,threeparttable,booktabs,mathtools,
dsfont,parskip,xr,tikz,times,w-thm}
\externaldocument[sm-]{supplement1}
\usetikzlibrary{bayesnet}

@@

  \title[Drug sensitivity prediction with the NIG prior]
  {Drug sensitivity prediction with the normal inverse Gaussian prior 
  informed by external data}

@@

	\subsection{Bayesian prior model}
	We carry out inference by endowing the parameters with the following priors:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), 
		  \label{eq:betaprior}\\
		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), \\
		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
		  \label{eq:tauprior}\\
		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
	with mean $\phi$ and shape $\lambda$. Prior distributions of the form
	(\ref{eq:prior}) are often referred to as global-local shrinkage rules 
	\cite[]{bernardo_shrink_2011}, due to the multiplicative separation of the 
	prior 
	variance into a local component $\gamma_{jd}^2$ and a global component
	$\tau_d^2$. The local components are supposed to capture local variation in
	the model parameters $\beta_{jd}$, while the global component is supposed to
	capture the general trend in $\bm{\beta}_{d}$. For appropriate 
	local shrinkage in global-local 
	shrinkage models it is important  to account for 
  different noise levels $\sigma_d^2$ by scaling the $\beta_{jd}$ variances 
  accordingly.

	The normal inverse Gaussian (NIG) prior model was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and since 
  \cite{barndorff-nielsen_normal_1997} it is routinely applied in mathematical 
  finance (see, e.g., \cite{kalemanova_normal_2007}). Here we extend it with an
  additional global variance component $\tau_d^2$. Supplementary Material
  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
  To illustrate the effect of the NIG prior on the posterior mean, we consider 
  the prior reparametrised as in 
  \cite{carvalho_handling_2009}, i.e., in terms of shrinkage weights 
  $\kappa_{jd}=1/(1 + \gamma_{jd}^2) \in (0, 1)$.
  Under the (simplified) normal means model, i.e., 
  $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
  $\tau_d^2=\sigma_d^2=1$, the resulting conditional
  posterior mean for the $\beta_{jd}$ is
  $\E(\beta_{jd} | y_{jd}, \kappa_{jd}) = (1 - \kappa_{jd}) y_{jd}$. 
  Thus, $\kappa_{jd}=0$ implies no shrinkage of $\beta_{jd}$ and $\kappa_{jd}=1$
  implies full shrinkage towards zero. Figure 
  \ref{fig:dens_kappa} depicts the prior on $\kappa_{jd}$ implied by 
  several choices of $\beta_{jd}$ prior.

<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors. Different line types correspond to different hyperparameter settings. The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
@

@@

  \subsection{Empirical Bayes}\label{sec:empiricalbayes}
  We parametrised the prior 
  mean of the $\gamma_{jd}^2$ as $\phi_{jd}=(\mathbf{c}_{jd} \tr 
  \bm{\alpha}_{\text{feat}})^{-1}$ and the prior mean of $\tau_d^2$ as
  $\chi_{d}=(\mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}})^{-1}$.
  This parametrisation allows us to include feature and drug covariates,
  both continuous and discrete, into the 
  model. Additionally, it reduces the number of hyperparameters from 
  $pD$ to $|\bm{\alpha}_{\text{feat}}| + |\bm{\alpha}_{\text{drug}}| + 2$. The
  Bayesian model
  then requires the specification of the hyperparameters 
  $\bm{\alpha}=\begin{bmatrix} \bm{\alpha}_{\text{feat}} \tr & 
  \bm{\alpha}_{\text{drug}} \tr 
  \end{bmatrix} \tr$ and $\bm{\lambda} = \begin{bmatrix} \lambda_{\text{feat}} &
  \lambda_{\text{drug}}
  \end{bmatrix} \tr$. These are
  abstract and hard to interpret parameters 
  for which we generally lack expert knowledge. They do, however, have a 
  significant influence on the shape of the posterior distribution. We 
  therefore propose to estimate these hyperparameters by empirical Bayes.
  In our case, this results in an   
  objective and data-driven inclusion of the feature and drug covariates.
	The canonical method for empirical Bayes is to maximise the marginal 
	likelihood with respect to the hyperparameters. In 
	\cite{casella_empirical_2001} the marginal likelihood is maximised by an EM 
	algorithm:
	\begin{align*}
	  \bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} & = \underset{\bm{\alpha},
	  \bm{\lambda}}{\argmax}\E_{\cdot | \Y} 
	  [\log p(\Y, \B, \bm{\Gamma}^2, \bm{\tau}^2, \bsigma^2) | \bm{\alpha}^{(l)},
	  \bm{\lambda}^{(l)}] \\
	  & = \underset{\bm{\alpha},\bm{\lambda}}{\argmax} 
	  \E_{\cdot | \Y} [\log \pi (\bm{\Gamma}^2) | \bm{\alpha}_{\text{feat}}^{(l)},
	  \bm{\lambda}_{\text{feat}}^{(l)}] + 
	  \E_{\cdot | \Y} [\log \pi (\bm{\tau}^2) | 
	  \bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
	\end{align*}
	where $\Y = \begin{bmatrix} \y_1 & \cdots & \y_D \end{bmatrix}$,
	$\B = \begin{bmatrix} \bbeta_1, \dots, \bbeta_D \end{bmatrix}$, 
	$\bm{\tau}^2 = \begin{bmatrix} \tau^2_1 & \cdots & 
	\tau^2_D \end{bmatrix} \tr$,
	$\bsigma^2 = \begin{bmatrix} \sigma^2_1 & \cdots & 
	\sigma^2_D \end{bmatrix} \tr$,
	and $\bm{\Gamma}^2 = \begin{bmatrix} \bgamma^2_1 & \cdots & \bgamma^2_D 
	\end{bmatrix}$,
	and the expectation is with respect to the joint posterior. In our case, this 
	posterior is not available in closed form, which renders the expectation 
	difficult. While \cite{casella_empirical_2001} suggests to approximate the 
	expectation by a Monte Carlo sample, we propose to use the variational Bayes 
	approximation developed in Section \ref{sec:variationalbayes}:
	$$
	\bm{\alpha}^{(l+1)},\bm{\lambda}^{(l+1)} = 
	\underset{\bm{\alpha},\bm{\lambda}}{\argmax}\E_{Q^{(l)}} 
	[\log \pi(\bm{\Gamma}^2)| \bm{\alpha}_{\text{feat}}^{(l)}, 
	\bm{\lambda}_{\text{feat}}^{(l)}] + 
	\E_{Q^{(l)}} [\log \pi(\bm{\tau}^2)| 
	\bm{\alpha}_{\text{drug}}^{(l)}, \bm{\lambda}_{\text{drug}}^{(l)}],
	$$
	where now the expectation is with respect to the converged variational 
	posterior $Q^{(l)}=\prod_{d=1}^D Q_d^{(l)}$.

@@

	\section{Simulations}\label{sec:simulations}
	\subsection{Setup}
	This section investigates the empirical 
	Bayes estimation properties of the model in a simulated setting; its main aim 
	is to assess hyperparameter recovery. It is a data based simulation,
	wherein the responses are simulated from a synthetic model, but the
	features are taken from the real GDSC data introducted in Section 
	\ref{sec:gdsc}. 
	The real GDSC features contain strong collinearities. Such strong
	collinearities in the design matrix impede correct parameter estimation. 
	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
	with the more modest aim of approximately correct estimation of the 
	hyperparameters.

	A pre-processing step selects 100 features with largest variance, while the 
	168 drug sensitivities for 482 cell lines are 
	simulated from three versions of model (\ref{eq:likelihood}) and 
	(\ref{eq:prior}). All versions draw the error variances as 
	$\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
	such that the prior $\sigma_d^2$ mean and variance are both one. 
	Version 1 fixes $\forall d: \tau_d^2=1$, version 2 fixes
	$\forall j,d: \gamma_{jd}^2=1$, while version 3 draws both the $\tau_d^2$ and
	$\gamma_{jd}^2$ according to (\ref{eq:prior}). In version 1, we
	create four external dummy feature covariates that code for four 
	randomly assigned approximately equally sized groups of features. We set
	$\bm{\alpha}_{\text{feat}}$ such
	that the $\gamma_{jd}^2$ of the four groups of features have prior means
	$\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$ 
	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. Version 2 
	follows a similar procedure for the $\tau_d^2$. We randomly create four
	groups of drugs with corresponding external drug dummy variables and set
	$\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$,
	such that we have $\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The
	scale is set to $\lambda_{\text{drug}}=1$. Version 3 combines the procedures
	from versions 1 and 2 to draw the $\gamma_{jd}^2$ and $\tau_d^2$.

	We estimate four models: (i) an NIG model that does include external 
	covariates in the estimation called NIG1 in the following, (ii) the NIG model
	estimated as in Section \ref{sec:estimation} called NIG2, and regular (iii)
	lasso and (iv) ridge models.
	Exclusion of the external covariates as in the NIG1 model
	amounts to direct estimation of one common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in the NIG2 model. All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines. We repeat the whole
	simulation 50 times.

	\subsection{Results}
<<simulation_gdsc_res1>>=
res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp
true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)
tab <- cbind(apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="elbot", ], 2, median) -
               apply(res[rownames(res)=="elbo", ], 2, median),
             apply(res[rownames(res)=="emse", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median),
             apply(res[rownames(res)=="pmse", ], 2, median) - 
               apply(res[rownames(res)=="pmset", ], 2, median))
tab <- round(tab, 3)
tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tab[which(tab[, 1]==max(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5] <- 
  paste0("\\textbf{", tab[which(tab[, 5]==min(tab[, 5], na.rm=TRUE)), 5], "}")
colnames(tab) <- c("$\\text{ELBO}_{\\text{test}}$",
                   "$\\text{ELBO}_{\\text{train}} 
                   - \\text{ELBO}_{\\text{test}}$",
                   "$\\text{EMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}} 
                   - \\text{PMSE}_{\\text{train}}$")     
@	
  Figure \ref{fig:simulations_gdsc_est1} shows the estimated
	$\bm{\alpha}_{\text{feat}}$ together with its true value for NIG2 in 
	version 1 of the simulation study. Figure 
	\ref{fig:simulations_gdsc_est1}a shows that 
	$\bm{\alpha}_{\text{feat}}$ is slightly but 
	consistently underestimated. On the $\phi_{jd}$ scale this underestimation is
	almost undetectable as depicted in Figure
	\ref{fig:simulations_gdsc_est1}b. $\lambda_{\text{feat}}$ is slightly 
	overestimated with median estimate \Sexpr{round(est.lambdaf[2], 2)} (the
	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
	a common phenomenon in variational Bayes approximations. In addition, 
	$\lambda_{\text{feat}}$ is a higher level scale
	parameter, which are, in general, hard to estimate. 
<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulation version 1 with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>= 
@

	Model NIG1 gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
	$\lambda_{\text{feat}}$ is slightly underestimated with median estimate
	\Sexpr{round(est.lambdaf[1], 2)} (the true value is one) and thus 
	overestimates the 
	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
	prior mean $\phi$, that is not able to capture part of the variation in the 
	true feature specific $\phi_{jd}$.

	Table \ref{tab:simulations_gdsc_tab1} displays several median performance
	measures for simulation version 1. The table includes as a reference 
	the null model, which
	is an empty model that only fits the means of the $\y_{d}$.
	The median evidence lower bounds (ELBO) approximates the marginal
	likelihoods of models NIG1 and NIG2 on both training and test data, while 
	estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
	performance. Focussing on MSE, we see that the NIG2 model, that includes
	the external covariates, performs best, both in terms of estimation and 
	prediction. Model fit is best for the NIG1 model that does not include the
	external covariates, according to the ELBO, calculated on the test data.
	The lower MSEs confirm that the model that
	includes the external data (NIG2) learns the underlying structure 
	in the data better than model NIG1 that does not include the external 
	data.
<<simulations_gdsc_tab1>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure estimates for simulation version 1 (best performing model in bold).", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@
